---
title: "Coding Notes"
description: |
  Coding Notes and Resources
author: 
  - name: Alex Stephenson
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
site: distill::distill_website
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Basic R Operations 

There are a voluminous amount of resources on R online. Here is a very minimal crash course. 

### Data Types 

R has six basic data types. The four most common types we work with are vectors, lists, matrices, and data frames. 

#### Vectors 

```{r}
## Three ways to create a vector
v = c(1,2,3,4,5)
v2 = seq(from=1,to=5, by = 1)

## Since the sequence is counting with no breaks we can also do this
v3 = 1:5
```

Vectors can also be characters (strings in other languages) 

```{r}
w = c("A", "B", "C")
w2 = c("1","2","3")

## We can convert to other types by the as.* series 
w3 = as.character(v)
w4 = as.numeric(w2)
```


#### Lists 

```{r}
l = list(v, w)
l
```

#### Matrices 

```{r}
m = matrix(seq(1,16,1), nrow = 4, byrow = T)
m
```

#### Data Frames 

```{r}
df = data.frame(
  var1 = letters[1:5],
  var2 = c(1,2,3,4,5)
)
df
```

### Subsetting 

### Control Operations 

### Functions

## Graphing in R 

## Simulations in R 

### The Birthday Problem 




